Pushdown Control-Flow Analysis via Refunctionalization

### 2018-02-23 paper.tex
update

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Abstracting abstract machines (AAM) is a systematic methodology for constructing abstract interpreter
  from a concrete small-step abstract machine. Recent progress applies the same idea on concrete definitional 
  interpreter and obtains a big-step abstracting definitional interpreter (ADI) written in monadic style.
  But the relation between small-step AAM and big-step ADI is not well-studied.

  % Mention Danvy's work
  In this paper, we show their correspondence and a series of transformations from a small-step 
  abstracting abstract machine to a big-step abstracting definitional interpreter. 
  The transformations include fusing, disentangling, 
  refunctionalizing and then monadification/un-CPS, each of them properly handle the collecting semantics and 
  the non-determinism of abstract interpretation. 
  The transformations also uncover several intermediary forms of abstract interpreter.
  Remarkably, we reveal how computatble precise call/return match in control-flow analysis is obtained by 
  refunctionalizing a small-step abstracting abstract machine with unbounded stack, 
  as well as explain how it been lost by defunctionalizing the higher-order continuations into 
  an algebraic data type.

  TODO: Shiver's formulation of CFA in denotational style.
  %Further, from another direction, we show Shiver's original implementation of control-flow analysis
  %in denotational style can be also transformed into natural semantics by closure conversion.

@@

\begin{itemize}
  \item We fill the gap between small-step AAM and big-step ADI by developing a series of systematic transformations from small-step AAM to big-step ADI, and also show that the reversed direction is also possible.

  \item We present an abstract interpreter written in CPS obtained from refunctionalization of small-step AAM. This abstract interpreter naturally has pushdown control-flow property, and we show how precise call/return match is obtained by representing evaluation context as a continuation function, as well as how it is lost by representing continuation as a data type in the defining language. We also analyze its termination, soundness and complexity property.

@@

@@

In this section, we first describe the syntax of A-Normal Form (ANF) $\lambda$-calculus and a CESK machine for executing it concretely. Then we review several recent works in control flow analysis: 1) Abstracting Abstract Machine as an abstract counterpart of CESK machine, 2) abstract interpreter written in a monadically-parameterized style, and 3) precise call/return match problem and existing solutions.

\subsection{A-Normal Form $\lambda$-Calculus}

@@

>>>>

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
%% http://ctan.org/pkg/subcaption
\usepackage{syntax}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Abstracting abstract machines (AAM) is a systematic methodology 
  for constructing abstract interpreter that derived from a concrete
  small-step abstract machine. Recent progress applies the same idea 
  on concrete definitional interpreter and obtains a big-step 
  abstracting definitional interpreter (ADI) written in monadic style.
  Danvy shows the correspondence between abstract machines and interpreters.
  In this paper, we study the relations between small-step AAM and 
  big-step ADI.

  We show their correspondence and how to syntactically transform a small-step 
  abstracting abstract machine to a big-step abstracting definitional 
  interpreter.
  The transformations include fusing, disentangling, refunctionalizing 
  and monadification(un-CPS), and all of them properly handle the 
  collecting semantics and the non-determinism of abstract interpretation. 
  After each transformation, we also obtain an intermediary form of 
  abstract interpreter.
  Remarkably, follow the idea that evaluation contexts are defuntionalized 
  continuations, we reveal how precise call/return match in control-flow 
  analysis is obtained by refunctionalizing a small-step abstracting 
  abstract machine,  % with unbounded stack
  as well as explain how it been lost by defunctionalizing the higher-order
  continuations into an algebraic data type. Further, we show how to turn 
  precise call/return match into computable analysis.

  TODO: Maybe something Shiver's formulation of CFA in denotational style.

@@

theoretical and practical

\begin{itemize}
  \item We fill the gap between small-step AAM and big-step ADI 
    by developing a series of systematic transformations from 
    small-step AAM to big-step ADI, and also the reversed 
    direction is also possible.

  \item We present an abstract interpreter written in continuation-passing
    style that is obtained from refunctionalization of a small-step AAM.
    This abstract interpreter naturally has pushdown control-flow property, 
    and we show how precise call/return match is obtained by 
    representing evaluation context as a continuation function, 
    as well as how it is lost by representing continuation as a 
    data type in the defining language.
    As a form of abstract interpreter, We also analyze its termination,
    soundness and complexity property.

@@

\subsection{Style}

We use Scala language to demonstrate the idea and each step of transformations.

@@

In this section, we frist describe the target language we will
analyze and a concrete CESK machine as operational semantics for it.
Then we review several recent works in control flow analysis for
higher-order languages:
1) abstracting abstract machines (AAM) as an abstract interpretation
counterpart of CESK machines,
2) definitional abstract interpreter written in a monadically-parameterized
style, and
3) precise call/return match problem in control-flow analysis and its existing solutions.
The abstracting abstract machines form the basis of our transformations, and
monadic abstract interpreter is the end point. In the meanwhile, the precise
call/return match is obtained during transformation.

\subsection{A-Normal Form $\lambda$-Calculus} \label{anfsyntax}

Tranditionally, continuation-passing style is a popular intermediate representation
for analyzing functional programs because of it exposes control transfer explicitly
and simplifies the analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use a direct-style intermediate representation as target for
clarity but without losing simplicty and generality.
The transformations we will show in the rest of this paper
also work on abstract machines for plain $\lambda$-calculus.

To begin, we now present the syntax of a call-by-value $\lambda$-calculus language in
administrative normal form (ANF) \cite{flanagan1993essence}.

\begin{verbatim}
e \in Exp ::= ae
            | (let ([x (f ae)]) e)
            | (letrec ([x ae] ...) e) (TODO: maybe not include letrec?)
ae \in AExp ::= x
              | lam
lam \in Lam ::= (lambda (x) e)
x ... are variable names
\end{verbatim}

In ANF, all the function applications must be administrated in a \texttt{let} expression,
and then bounded to a variable name under current environment.
We can use \texttt{letrec} to create mutually recursive function bindings. But \texttt{letrec}
only accepts atomic expressions that appear at the right-hand side of a binding.
The atomic expressions $ae$ are either a variable, or a literal \texttt{lambda} term, which
both can be evaluated in a single step.

The abstract syntax tree can be straightforwardly described in Scala as follows:

\begin{verbatim}
sealed trait Expr
case class Binding(x: String, e: Expr)

case class Var(x: String) extends Expr
case class App(e1: Expr, e2: Expr) extends Expr
case class Lam(x: String, body: Expr) extends Expr
case class Let(x: String, e: App, body: Expr) extends Expr
case class Letrec(bds: List[Binding], body: Expr) extends Expr
\end{verbatim}

@@

\textit{C}ontrol \textit{E}nvionment \textit{S}tore \textit{K}ontinuation (CESK) machine is an
abstract machine for describing semantics of and evaluating $\lambda$-calculus. (TODO: cite)
CESK machine models program execution as state transitions in small-step fashion. As its name suggested,
a machine state has four components: 1) Control is the expression currently being evaluated.
2) Environment is the a map that contains the address of a variable that in the lexical
scope of control.
3) Store models the heap of a program, and it is also a map but maps from addresses to values.
The addresses are infinite numbers starting from 0.
In our toy language, the only catagory of value is closure, i.e., a function paired with an environment.
4) Continuation represents the program stack.

We show the Scala representations for the components of CESK machine.

\begin{verbatim}
type Addr = Int
type Env = Map[String, Addr]
type Store = Map[Addr, Storable]

abstract class Storable
case class Clos(v: Lam, env: Env) extends Storable

case class Frame(x: String, e: Expr, env: Env)
type Kont = List[Frame]

case class State(e: Expr, env: Env, store: Store, k: Kont)
\end{verbatim}

It is worth noting that the continuation class \texttt{Kont} is defined as a list of frames.
The frame class \texttt{Frame} stores the information of call-site, i.e., the information that
used to resume the interrupted computation.
A \texttt{Frame} constitutes of a variable name \texttt{x} to be bind later, a control expression
that resumes to, and its environment.

Before go into describing how the machine evaluates expressions, we firstly define several helper functions.
As we mentioned in section \ref{anfsyntax}, the atomic expressions are either a variable, or
a literal \texttt{lambda} term. So the helper function \texttt{atomicEval} handles these two
cases and evaluates atomic expressions to closures in a straightforward way.
The \texttt{alloc} function generates a fresh address, and always allocates an unique integer under the domain
of store.
The \texttt{isAtomic} function is used as a predicate of whether the expression is atomic.

\begin{verbatim}
def atomicEval(e: Expr, env: Env, store: Store): Storable = e match {
  case Var(x) => store(env(x))
  case lam@Lam(x, body) => Clos(lam, env)
}
def alloc(store: Store): Addr = store.keys.size + 1
def isAtomic(e: Expr): Boolean = e.isInstanceOf[Var] || e.isInstanceOf[Lam]
\end{verbatim}

Now, we can faithfully describe the state transition function \texttt{step},
that given a machine state, function \texttt{step} determines its successor state.

\begin{verbatim}
def step(s: State): State = s match {
  case State(Let(x, App(f, ae), e), env, store, k) if isAtomic(ae) =>
    val Clos(Lam(v, body), env_c) = atomicEval(f, env, store)
    val addr = alloc(store)
    val new_env = env_c + (v -> addr)
    val new_store = store + (addr -> atomicEval(ae, env, store))
    val frame = Frame(x, e, env)
    State(body, new_env, new_store, frame::k)
  case State(ae, env, store, k) if isAtomic(ae) =>
    val Frame(x, e, env_k)::ks = k
    val addr = alloc(store)
    val new_env = env_k + (x -> addr)
    val new_store = store + (addr -> atomicEval(ae, env, store))
    State(e, new_env, new_store, ks)
}
\end{verbatim}

The first case is that the control of current state is a \texttt{Let} expression,
and its right-hand side is an function application.
By calling function \texttt{atomicEval}, we obtain the closure that callee \texttt{f} stands for.
The successor state's control then transfers to the \texttt{body} expression of closure,
with an updated environment and an update store. The new environment is extended
from closure's environment and mapped \texttt{v} to a fresh address \texttt{addr}.
The new store also extends with \texttt{addr} mapping to the value of \texttt{ae},
which is evaluated from \texttt{atomicEval}.
In the meanwhile, a new frame \texttt{frame} is pushed onto the stack \texttt{k}.
This \texttt{frame} contains the left-hand side variable name of \texttt{Let},
the body expression, and the lexical environment of body expression.

The second case for \texttt{step} is the control is transferred to an atomic expression.
We extract the top frame of continuations firstly.
The control (that is an atomic expression) of current state is the evaluated term
that being bind to the variable \texttt{x} from the top frame.
The environment and store are updated with \texttt{x} mapping to the closure value of \texttt{ae}.
Then the successor state is transferred to expression \texttt{e} from the top frame,
which is the body of a \texttt{Let} expression, with the updated environment, store, and
the rest of stack \texttt{ks}.

To run the program,

\begin{verbatim}
def drive(s: State): State = s match {
  case State(ae, _, _, Nil) if isAtomic(ae) => s
  case _ => drive(step(s))
}

def inject(e: Expr): State = State(e, Map(), Map(), Nil)
def eval(e: Expr): State = drive(inject(e))
\end{verbatim}
